Leadership across cultures and contexts: an exploratory analysis

### 2020-05-09 leadership_across_cultures_contexts.Rmd
Working on discussion section

highstatus_plot <- highstatus_plot + scale_x_log10()
highstatus_plot + plot_elastic_status + plot_annotation(title= "Predictors of High status", tag_levels = 'A', caption = 'Odds ratio', theme=theme(plot.caption=element_text(hjust=0.5, size=12, margin=margin(t=0))))
```

@@

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in over 35% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by subsistence strategy with hunter-gatherers providing relatively higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

coercive_plot_plot <- coercive_plot + scale_x_log10()
coercive_plot_plot + plot_elastic_coercive + plot_annotation(title= "Predictors of Coercive authority", tag_levels = 'A', caption = 'Odds ratio', theme=theme(plot.caption=element_text(hjust=0.5, size=12, margin=margin(t=0))))
```

@@

The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural pressures, and receiving various benefits in exchange for their functions. See Figure \@ref(fig:featuressupportplot).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social material success" feature, which included Polygyny, High quality spouse, and Many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for Social material success was higher for females leaders than male leaders (see Figure \@ref(fig:heatmap_feature_sex)), with the caveat that there were few text records on female leaders in our sample.

@@

The distribution of text records by group context and subsistence strategy suggests that although leaders emerge within similar social contexts across populations, it occurs at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, despite potential for residential mobility, do rely in part on a highly defensible resource base (i.e., livestock) and exhibit political leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992; @glowacki_evolutionary_2017]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among nonindustrial agricultural populations which are characterized by political leadership systems beyond the community level -- more similar to political structures of WEIRD populations -- the importance of leadership within residential and family groups persists. For example, among rural Korean populations a chairwomen of local 'club' organizations maintains an important, although informal, leadership position within the group and represents the group to governmental bodies. @aa01-062 [p. 173] describes,

@@

Several theoretically important leadership dimensions were relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The dominance theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustrial cultures sampled here. Clearly there is substantial diversity in many dimensions of leadership among non-WEIRD populations.

>>>>

highstatus_plot <- highstatus_plot + scale_x_log10()
highstatus_plot + plot_elastic_status + 
  plot_annotation(
    title = 'Predictors of High status',
    tag_levels = 'A', 
    caption = 'Odds ratio', 
    theme=theme(plot.caption=element_text(hjust=0.5, size=12, margin=margin(t=0)))
    )

@@

```{r}
nocoerciveauthority <- signif(100*qualities_support_cult$Estimate[qualities_support_cult$Variable == 'No coercive authority'], 3)
```

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in `r nocoerciveauthority`% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by subsistence strategy with hunter-gatherers providing relatively higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

coercive_plot <- coercive_plot + scale_x_log10()
coercive_plot + plot_elastic_coercive +
  plot_annotation(
    title = 'Predictors of Coercive authority',
    tag_levels = 'A', 
    caption = 'Odds ratio', 
    theme=theme(plot.caption=element_text(hjust=0.5, size=12, margin=margin(t=0)))
    )

@@

The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural norms, and receiving various benefits in exchange for their functions. See Figure \@ref(fig:featuressupportplot).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social material success" feature, which included Polygyny, High quality spouse, and Many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for Social material success was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

<!-- NEED TO DO: ADD SOME REFS AND BRIEF INTERPRETATION FOR BELOW-->

Four leadership dimensions varied by subsistence strategy. There was more evidence that hunter-gatherer leaders lacked coercive authority compared to other leaders. There was more evidence that both hunter-gatherer and pastoralist leaders generated food surpluses, but less evidence that pastoralist leaders distributed food and other resources.  There was more evidence that leaders in communities with mixed subsistence strategies distributed resources and less that they were older.

The distribution of text records by group context and subsistence strategy (see Figure \@ref(fig:grouptype)) suggests that although leaders emerge within similar social contexts across populations, they do so at variable frequencies. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, despite potential for residential mobility, do rely in part on a highly defensible resource base (i.e., livestock) and exhibit evidence for political leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992; @glowacki_evolutionary_2017]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between evidence for political leadership at the community level or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among nonindustrial agricultural populations which are characterized by political leadership systems beyond the community level -- more similar to political structures of WEIRD populations -- the evidence of leadership within residential and family groups persists. For example, among rural Korean populations a chairwomen of local 'club' organizations maintains an important, although informal, leadership position within the group and represents the group to governmental bodies. @aa01-062 [p. 173] describes,

@@

Evidence for several theoretically important leadership dimensions was relatively rare in the ethnographic record. The nonhuman animal leadership literature, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet evidence for leaders as sources of moral authority or "fair" was found in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The dominance theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures was there evidence that leaders were feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but evidence that leaders were attractive was found in only `r attractive_pct`% of the nonindustrial cultures sampled here. Clearly there is substantial diversity in many dimensions of leadership among non-WEIRD populations.
